|

### 2021-08-09 prova_knit_finale.Rmd
convergence

\section{Convergence}\label{pf_converg}

>>>>

\section{Convergence}\label{pf_converg}
In this section, we cover three propositions that justify the use of the PF algorithms to estimate quantities of interest. Firstly, we present two convergence results for the estimators of the predictive and filtering distributions. As we will discuss, these results hold under reasonable conditions. Then, we will show that the error of the particle estimators are asymptotically distributed according to a Gaussian distribution with known variance. 
Henceforth, we will maintain the assumption that (multinomial) resampling is repeated at each iteration.
\subsection{MSE convergence}
As already seen in subsection \ref{seq_impt_sampling}, the particle approximation of $\Q_{t}(dx_{0:t})$ is:
\begin{equation*}
      \sum_{n=1}^NW_{t-1}^nM_t(X_{t-1}^n,dx_t).
\end{equation*}
The following lemma states that the (mean squared) error related to sampling from this conditional distribution is bounded.
\bigskip
\begin{lemma}
For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
$$
\E\Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg\}^2 \Bigg]\leq \frac{\|\varphi\|_{\infty}^2}{N},
$$
where $\|\cdot\|_{\infty}$ is the supnorm and $M_t\varphi$ is the map 
$x_{t-1}\mapsto \int_{\X}M_t(x_{t-1},dx_t)\varphi(x_t).$
\end{lemma}
This result can be used to prove the convergence of the estimator $\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)$. To see this, note that 
\begin{equation}
    \begin{split}
        \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)=\bigg(&\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg) \\
        & + \bigg(\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n)- \E_{\Q_{t-1}M_t}(\varphi)\bigg).
    \end{split}
\end{equation}
where $\E_{\Q_{t-1}M_t}(\varphi)=\int_{\X^{t+1}}\varphi(x_t)  \Q_{t-1}(dx_{0:t-1})M_t(x_{t-1},dx_t)$.\
By the lemma, the first component of the right-hand side has a bounded MSE. It can be proved that also the second component of the right-hand side has a bounded MSE. Using the fact that $\E\{(X+Y)^2\}\leq2\E\{\E(X^2)+\E(Y^2)\},$ it follows that also the MSE of the left-hand side is bounded. A simple induction argument proves that the result holds for all $t$.\
Using a similar argument, one can prove that the MSE of the estimator $\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)$ is bounded as well. These ideas are formalized in the following proposition.
\bigskip
\begin{proposition}
For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$ $\forall t\geq0$ there exist constants $c_t,c_t'$ such that:
$$
\E \Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg\}^2 \Bigg]\leq c_t \frac{\|\varphi\|_{\infty}^2}{N}
$$
and 
$$
    \E \Bigg[ \bigg\{\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_t}(\varphi) \bigg\}^2 \Bigg]\leq c_t' \frac{\|\varphi\|_{\infty}^2}{N}.
$$
Therefore, taking the limit for $N\rightarrow \infty,$ we have that 
$$
\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_{t-1}M_t}(\varphi)
$$
and 
$$
\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_t}(\varphi)
$$
\end{proposition}
In the next subsection, we establish a much stronger property of the particle filter algorithms.

\subsection{Almost sure convergence}
The following proposition states that estimators $\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)$ and $\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)$ converge to their true values, up to negligible sets, i.e., the sets such that the estimators do not converge to $\E_{\Q_{t-1}M_t}(\varphi)$ and $\E_{\Q_t}(\varphi)$ have zero probability.  Additionally, recall that almost sure convergence implies convergence in probability.
\bigskip
\begin{proposition}
For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
$$
\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{a.s}}\E_{\Q_{t-1}M_t}(\varphi)
$$
and 
$$
\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{a.s}}\E_{\Q_{t}}(\varphi).
$$
\end{proposition}

\subsection{Central limit theorem}
The last asymptotic result of PF estimators concerns convergence in law: the error of a particle estimate is asymptotically normally distributed. 
\bigskip
\begin{proposition}
For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
$$
\sqrt{N}\bigg( \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg)\implies \N(0,\hat{\mathcal{V}}_t(\varphi))
$$
and 
$$
\sqrt{N}\bigg(\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_{t}}(\varphi)\bigg)\implies \N(0,\mathcal{V}_t(\varphi)),
$$
where the asymptotic variances are defined recursively, with $ \hat{\mathcal{V}}_0(\varphi):=Var_{\M_0}(\varphi) $,
\begin{equation*}
    \begin{split}
        \hat{\mathcal{V}}_t(\varphi)&:=\mathcal{V}_{t-1}(\E_{M_t}(\varphi))+Var_{\Q_{t-1}M_t}(\varphi) \text{ }\forall t\geq1, \\
        \mathcal{V}_t(\varphi)&:=\hat{\mathcal{V}}_{t}\big( \bar{G}_t(\varphi-\E_{\Q_t}(\varphi)) \big) \text{ }\forall t\geq0.
    \end{split}
\end{equation*}
Here, $\bar{G}_t=G_t/\E_{\Q_{t-1}M_t}(G_t)$ is the normalized incremental weight.
\end{proposition}
This version of the CLT for PF suggest that PF estimators with respect both to the predictive and the filtering distributions are approximately normal for a large number of particles. \
In particular, the logic behind the expression for $\mathcal{V}_t(\varphi)$ is the same as described in section \ref{asympt_IS}. In fact, recall that the estimator with respect to the filtering distribution is a more sophisticated form of IS estimator, where the target distribution differs from the proposal one. \ 
Instead, to get an intuition of the variance $\hat{\mathcal{V}}_t(\varphi),$ note that by the law of total variance, 
\begin{equation*}
    \begin{split}
        Var\Bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n) \Bigg] = &Var\Bigg[ \E\bigg[\frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)\Big| \sigma(X_{t-1}^{1:N}) \bigg] \Bigg] \\
        &+ \E \Bigg[ Var\bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)    \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg] \\
        &= Var\bigg[\sum\limits_{n=1}^N W_{t-1}^n (M_t\varphi)(X_{t-1}^n)\bigg] \\
        &+\frac{1}{N}\E\Bigg[Var\bigg[\varphi(X_t^1) \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg].
    \end{split}
\end{equation*}
The notation $\mid \sigma(X_{t-1}^{1:N})$ means conditioning on having sampled (from the multinomial distribution) $X_{t-1}^{1:N}$. Conditional on multinomial sampling, the $X_{t-1}^n$ are i.i.d., with distribution  $\sum\limits_{n=1}^N W_{t-1}^n M_t(X_{t-1}^n, dx_t)$, that should converge to $\Q_{t-1}M_t$. So, it is natural to think that the second element of the sum converges to the variance of $\varphi$ with respect to the distribution $\Q_{t-1}M_t$. 
These results are of practical use because they provide a way to explicitly compute the variance of the PF estimates.
